---
title: "Filtering cells"
author:
    - name: "[Ciro Ramirez Suastegui](https://github.com/cramirezs)"
      affiliation: "La Jolla Institute for Immunology, San Diego, USA"
      email: "ksuasteguic@gmail.com"
date: "Last update: `r format(Sys.time(), '%Y, %B %d, at %H:%M')`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    number_sections: false
    fig_caption: true
    theme: cosmo
    code_folding: hide
---

<!--
R -e "rmarkdown::render(
  input = '/home/ciro/scripts/clustering/R/filtering_assessment.rmd',
  output_dir = '/home/ciro/ad_hoc/pbtumor/results/'
)" --slave
-->


This script is to address the filtering differences between:

- `/home/ciro/scripts/quality_control/single_cell.R`
- `/home/ciro/scripts/clustering/R/seurat.R`


We are taking the [PBT project](https://informaticsdata.liai.org/NGS_analyses/ad_hoc/Groups/vd-vijay/cramirez/pbtumor/results/CD45pCD3p_report.html) as example.
The missing doublet filter won't affect the assessment

```{bash, parameters_in_config}
grep -EC 2 "_RNA|percent.mt|HT_ID|nSamples_expressed" /home/ciro/pbtumor/scripts/qc_config_CD45pCD3p.yaml
grep -EC 2 "nSamples_expressed" /home/ciro/scripts/quality_control/single_cell.R
grep -EC 2 "_RNA|percent.mt|HT_ID|nSamples_expressed" /home/ciro/ad_hoc/pbtumor/results/clustering/CD45pCD3p_xdoublets/config.yaml
```

### Correlation between metrics

```{r, metadata_cor}
source("/home/ciro/scripts/handy_functions/devel/file_reading.R")
mdata_pre = readfile("/home/ciro/ad_hoc/pbtumor/results/quality_control/CD45pCD3p/metadata_filtered.rdata")
mdata_pos = readfile(paste0("/home/ciro/ad_hoc/pbtumor/results/clustering/",
  "CD45pCD3p_xdoublets/.object_meta.data_seurat_mean0.01_pct15_pc15.rds"))
cnames = c("nCount_RNA", "nFeature_RNA", "percent.mt")
sapply(X = cnames, FUN = function(x){
  cor(mdata_pre[rownames(mdata_pos), x], mdata_pos[, x])
})
```

You can see that the correlation is good.

### 1. quality_control/single_cell.R

```{bash, code_qc}
grep 'Minimal filter:' -A 100 /home/ciro/scripts/quality_control/single_cell.R |
  grep -EB 100 'Data = casefold|RNA|meta_data <- meta_data|expr_data <- expr_data'
```

1.1 Filter cells with the low nFeature_RNA (max: 200).

1.2 Filter features expressed in a percentage of the cells (min: 0, max: 30 cells).

1.3 Calculate percent.mt[ribo,hs] and _RNA metrics.
Another point to note is that we could make filtering by the metadata the
very first step, but that would make the metrics and number of features
change depending on the subset and I am not sure if that's something we want.
`grep -EC 3 "subsampling" /home/ciro/scripts/quality_control/single_cell.R`


### 2. CreateSeuratObject recalculates _RNA metrics

```{bash, code_clustering}
grep 'Filtering data' -A 50 /home/ciro/scripts/clustering/R/seurat.R |
  grep -EB 50 '\);|nSamples_min|expr_data'
```

2.1 Filter metadata based on filtering$subset (YAML).

2.2 Filter features expressed in a percentage of the cells
(calculated now with filtered data).

2.3 The _RNA metrics values are now from CreateSeuratObject.

### Conclusion

The _RNA metrics and number of features we end up with changes by using
CreateSeuratObject after filtering the metadata. Also,
quality_control/single_cell.R is constrained by wanting a "Minimal filter."
Firstly, this excludes cells with features dictated by the lower bound (0 and 200).
Secondly, "nSamples_expressed" dictates the amount of cells a feature needs to
be expressed in (this will be between 0 and 30 cells).
Only the second happens just before the clustering. Thus varying the amount of
features the clustering takes.

Alternatives:

1. Testing if CreateSeuratObject would come up with the same numbers by
repeating the same "Minimal filters."

2. Don't let CreateSeuratObject estimate the metrics.

I believe the numbers are close enough to not affect. But this needs to be
assessed with more extreme filters. In any case, alternative (2) should not be
used because these metrics need to vary depending on the subset.

---

```{r report_progress, echo = FALSE, results = 'asis'}
cat("Report compiled from:", knitr::current_input(dir = TRUE), "\n")
```
